---
title: "Distance Sampling Analysis - Single Observer"
author: "Your name"
---

This is a template for the analysis of distance sampling data.

```{r, echo=FALSE}
suppressPackageStartupMessages(library(pander))
suppressPackageStartupMessages(library(mrds))
chitable <- function(observed, expected) {
	x = rbind(observed, expected, (observed - expected)^2/expected)
	x = cbind(x, apply(x, 1, sum))
    x=rbind(x[1,,drop=FALSE],t(apply(x[2:3,,drop=FALSE],1,function(x) sprintf("%0.3f",x))))
	colnames(x)[ncol(x)]="Total"
	x=rbind(bin=colnames(x),x)
	rownames(x) = c("Bin","Observed", "Expected", "Chisquare")
	return(x)
}
df <- read.csv("data.csv",header=TRUE)
```

Fit selected model and set bins used for display and chi-square distance intervals.
```{r}
model <- ddf_string
bins <-  bin_string
```

This is a summary of the fitted model results and parameters.

```{r, echo=FALSE}
print(model)
```

This is a plot of the fitted detection function as a line, predictions for individual observations as points and a histogram as an alternate description for the shape of the data.

```{r, echo=FALSE}
plot(model,breaks=bins)
```



This is a q-q plot of the goodness-of-fit for the fitted detection function.     
```{r, echo=FALSE}
result <- ddf.gof(model,breaks=bins)
if(!is.na(result$chisquare$chi1$p))
{
   chip <- paste("For the chi-square test, P =",sprintf("%0.3f",result$chisquare$chi1$p),
              " with ",result$chisquare$chi1$df," degrees of freedom",sep="")
} else{
   chip <- "No p-value can be constructed because there are no degrees of freedom for the test"
}
chi.table=t(chitable(result$chisquare$chi1$observed, result$chisquare$chi1$expected))
rownames(chi.table)=NULL
```

This is the chi-square table using the selected bins.
```{r, results='asis',echo=FALSE}
pander(chi.table)
```

`r chip `. The Kolmogorov-Smirnov test statistic is `r sprintf("%.3f",result$dsgof$ks$Dn,digits=3)` (P = `r sprintf("%.3f",result$dsgof$ks$p)`). The Cramer-von Mises test statistic(unweighted) is `r sprintf("%.3f",result$dsgof$CvM$W) ` (P = `r sprintf("%.3f",result$dsgof$CvM$p)`).


